---
title: "FinnPRIO — Effect of Management and Import Restrictions on Risk"
subtitle: "Phytosanitary measures (ENT2A vs ENT2B) versus import volume (ENT3)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
params:
  db_path: "databases/master_database/finnprio_master.db"
  zero_tol: 0.000001
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 110)

# Resolve all relative paths (db_path, CSV exports) from the project root so the
# report behaves the same whether knit from the Rmd folder or the project root.
# Applies from the NEXT chunk onward.
.input  <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NA)
.root   <- if (!is.na(.input)) {
  normalizePath(file.path(dirname(.input), "..", ".."))
} else {
  normalizePath(getwd())
}
knitr::opts_knit$set(root.dir = .root)

library(DBI)
library(RSQLite)
library(tidyverse)
library(jsonlite)
library(DT)
library(patchwork)

is_zero <- function(x) !is.na(x) & x <= params$zero_tol

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  plot.title = element_text(face = "bold")))
```

```{r connect, include=FALSE}
# ---- READ-ONLY connection -------------------------------------------------
# Opened with SQLITE_RO so this report can never write to the master database.
db_path <- params$db_path
stopifnot(file.exists(db_path))
con <- dbConnect(RSQLite::SQLite(), db_path, flags = RSQLite::SQLITE_RO)
knitr::knit_hooks$set(document = function(x) { try(dbDisconnect(con), silent = TRUE); x })
dir.create("scripts/exploration/output", showWarnings = FALSE, recursive = TRUE)

pal_shift <- c("No change"     = "#9e9e9e",
               "1 step lower"  = "#4daf4a",
               "≥2 steps lower"= "#1b7837",
               "Higher (invalid)" = "#d73027")
```

```{r load-data}
# ---- Raw tables -----------------------------------------------------------
assessments <- dbReadTable(con, "assessments")
pests       <- dbReadTable(con, "pests")
sims        <- dbReadTable(con, "simulations")
simsum      <- dbReadTable(con, "simulationSummaries")
pwAnswers   <- dbReadTable(con, "pathwayAnswers")
entryPath   <- dbReadTable(con, "entryPathways")
pathways    <- dbReadTable(con, "pathways")
pwQuestions <- dbReadTable(con, "pathwayQuestions")
questions   <- dbReadTable(con, "questions")
answers     <- dbReadTable(con, "answers")
taxa        <- dbReadTable(con, "taxonomicGroups")

# ---- Option -> points lookups (parsed from the `list` JSON) ---------------
parse_points <- function(df, id_col) {
  map_dfr(seq_len(nrow(df)), function(i) {
    lst <- jsonlite::fromJSON(df[["list"]][i])
    tibble(qid = df[[id_col]][i],
           opt = as.character(lst$opt),
           points = as.numeric(lst$points))
  })
}
pw_points <- parse_points(pwQuestions, "idPathQuestion")   # ENT2A=1, ENT2B=2, ENT3=3, ENT4=4
q_points  <- parse_points(questions,   "idQuestion")

# ---- Latest simulation per assessment -------------------------------------
latest_sim <- sims %>%
  group_by(idAssessment) %>%
  slice_max(idSimulation, n = 1, with_ties = FALSE) %>%
  ungroup()

# ---- Wide risk variables (medians + 5/95% bounds) -------------------------
risk_wide <- simsum %>%
  filter(idSimulation %in% latest_sim$idSimulation) %>%
  select(idSimulation, variable, median, q5, q95) %>%
  pivot_wider(names_from = variable, values_from = c(median, q5, q95))

# ---- Species-level table --------------------------------------------------
tax_lookup <- taxa %>% rename(idTaxa = 1, taxGroup = 2)

species <- latest_sim %>%
  select(idAssessment, idSimulation) %>%
  left_join(assessments %>% select(idAssessment, idPest, idAssessor,
                                   finished, valid), by = "idAssessment") %>%
  left_join(pests %>% select(idPest, scientificName, eppoCode, idTaxa), by = "idPest") %>%
  left_join(tax_lookup %>% select(idTaxa, taxGroup), by = "idTaxa") %>%
  left_join(risk_wide, by = "idSimulation") %>%
  mutate(taxGroup = ifelse(is.na(taxGroup), "Unknown", taxGroup))

n_species <- nrow(species)
```

```{r answer-prep}
# Attach points to each pathway answer for ENT2A/ENT2B/ENT3/ENT4.
# pwAnswers store the option code (a,b,c,...) in min/likely/max.
pw_long <- pwAnswers %>%
  select(idEntryPathway, idPathQuestion, min, likely, max) %>%
  left_join(entryPath %>% select(idEntryPathway, idAssessment, idPathway),
            by = "idEntryPathway") %>%
  left_join(pathways %>% select(idPathway, pathName = name, group), by = "idPathway") %>%
  # only keep the simulated assessments
  filter(idAssessment %in% species$idAssessment)

points_for <- function(qid, opt_vec) {
  lk <- pw_points %>% filter(qid == !!qid) %>% select(opt, points)
  tibble(opt = as.character(opt_vec)) %>%
    left_join(lk, by = "opt") %>% pull(points)
}

# ENT2A (qid 1) vs ENT2B (qid 2) on the `likely` answer, per pathway
ent2 <- pw_long %>%
  filter(idPathQuestion %in% c(1, 2)) %>%
  select(idEntryPathway, idAssessment, idPathway, pathName, group,
         idPathQuestion, likely) %>%
  pivot_wider(names_from = idPathQuestion, values_from = likely,
              names_prefix = "q") %>%
  rename(ent2a_opt = q1, ent2b_opt = q2) %>%
  mutate(ent2a_pts = points_for(1, ent2a_opt),
         ent2b_pts = points_for(2, ent2b_opt),
         delta = ent2b_pts - ent2a_pts,
         shift = case_when(
           is.na(delta)            ~ NA_character_,
           delta > 0               ~ "Higher (invalid)",
           delta == 0              ~ "No change",
           ent2a_pts - ent2b_pts <= 1 ~ "1 step lower",
           TRUE                    ~ "≥2 steps lower"))

# ENT3 (qid 3) import volume on the `likely` answer, per pathway
ent3 <- pw_long %>%
  filter(idPathQuestion == 3) %>%
  select(idEntryPathway, idAssessment, idPathway, group, likely) %>%
  mutate(ent3_pts = points_for(3, likely),
         ent3_cat = factor(recode(as.character(likely),
                                  a = "Non-existent", b = "Small",
                                  c = "Medium", d = "Large"),
                            levels = c("Non-existent","Small","Medium","Large")))

# Species-level import volume: highest import category across pathways
species_import <- ent3 %>%
  group_by(idAssessment) %>%
  summarise(max_import_pts = if (all(is.na(ent3_pts))) NA_real_ else max(ent3_pts, na.rm = TRUE),
            n_pathways = n(),
            n_nonexistent = sum(ent3_pts == 0, na.rm = TRUE),
            n_answered = sum(!is.na(ent3_pts)),
            .groups = "drop") %>%
  mutate(zero_import = n_answered > 0 & n_nonexistent == n_answered,
         max_import_cat = factor(recode(as.character(max_import_pts),
                                        `0` = "Non-existent", `1` = "Small",
                                        `2` = "Medium", `3` = "Large"),
                                 levels = c("Non-existent","Small","Medium","Large")))

species <- species %>% left_join(species_import, by = "idAssessment")
```

```{r headline-metrics}
# Management (A vs B) effect on risk, per species
mgmt <- species %>%
  transmute(idAssessment, scientificName, taxGroup, valid, finished,
            riskA = median_RISKA, riskB = median_RISKB,
            entryA = median_ENTRYA, entryB = median_ENTRYB,
            establishment = median_ESTABLISHMENT, impact = median_IMPACT,
            abs_reduction = riskA - riskB,
            rel_reduction = ifelse(riskA > params$zero_tol,
                                   (riskA - riskB) / riskA, NA_real_))

n_no_mgmt_effect <- sum(abs(mgmt$abs_reduction) <= params$zero_tol, na.rm = TRUE)
pct_no_effect    <- n_no_mgmt_effect / nrow(mgmt)
# Effect among species where measures actually change risk (RISK A > 0 and A != B)
affected <- mgmt %>% filter(riskA > params$zero_tol, abs_reduction > params$zero_tol)
median_rel_affected <- median(affected$rel_reduction, na.rm = TRUE)
n_affected   <- nrow(affected)
n_zero_import    <- sum(species$zero_import, na.rm = TRUE)
# Scenario B (with current measures) is the operative real-world risk;
# A (no measures) is the counterfactual used only to size the management effect.
n_zero_riskB     <- sum(is_zero(species$median_RISKB))   # actual: with measures
n_zero_riskA     <- sum(is_zero(species$median_RISKA))   # counterfactual: no measures
n_pushed_to_zero <- n_zero_riskB - n_zero_riskA          # extra species zeroed by measures
```

<style>
#report-sections > h1 { display: none; }   /* hide the tabset wrapper heading */
.nav-pills { margin-bottom: 1.4em; }
.tab-pane { padding-top: 1.4em; }
.tab-content > .tab-pane h3 { margin-top: 1.8em; }
.tab-content > .tab-pane > p { margin-bottom: 1.1em; }
blockquote { margin: 1.4em 0; }
</style>

# Report sections {.tabset .tabset-pills .tabset-fade}

## 1. Headline findings

This report uses the **latest stored simulation** for each of the
**`r n_species`** assessments in `finnprio_master.db` that have been simulated.
All figures are observational — read directly from the saved Monte-Carlo
summaries and the assessors' answer codes. No re-simulation is performed and the
database is opened read-only.

> **Phytosanitary measures leave most species unchanged — but are not weak
> where they apply.**
> Comparing scenario A (`ENT2A`, *no* official measures) with scenario B
> (`ENT2B`, *with* measures), the median risk is **identical** for
> **`r n_no_mgmt_effect`** of `r n_species` species
> (`r scales::percent(pct_no_effect, accuracy = 1)`) — the assessors gave the
> same `ENT2A` and `ENT2B` answers. For the **`r n_affected`** species where the
> answers *do* differ, however, risk falls by a median of
> **`r scales::percent(median_rel_affected, accuracy = 1)`**. Because `ENT2` is a
> *multiplicative* factor of entry, even a one-category answer change — which can
> look minor to an assessor — roughly halves risk. "Almost no difference in the
> answers" is therefore not the same as "almost no difference in risk".
>
> **No import sets risk to zero.**
> **`r n_zero_import`** species have *Non-existent* import volume (`ENT3`) on
> every entry pathway. Under current measures (scenario **B**, the operative
> risk) **`r n_zero_riskB`** species sit at median risk 0; without measures
> (scenario **A**) it would be **`r n_zero_riskA`** — so the measures push only
> **`r n_pushed_to_zero`** additional species to zero. Section 4 shows how much of
> the zero-risk set is driven by no-import versus establishment or impact.
>
> *Throughout, scenario **B** (with current official measures) is the default
> "actual" risk; scenario **A** appears only to quantify the management effect.*

```{r headline-table}
mgmt %>%
  summarise(
    `Species (simulated)`             = n(),
    `Median risk A (no measures)`     = round(median(riskA, na.rm = TRUE), 4),
    `Median risk B (with measures)`   = round(median(riskB, na.rm = TRUE), 4),
    `Species unchanged by measures`   = paste0(n_no_mgmt_effect, " (",
                                          scales::percent(pct_no_effect, accuracy = 1), ")"),
    `Species where measures differ`   = n_affected,
    `Median reduction where differ`   = scales::percent(median_rel_affected, accuracy = 1),
    `Zero-import species`             = n_zero_import,
    `Species at risk 0 (B, actual)`   = n_zero_riskB,
    `Species at risk 0 (A, no measures)` = n_zero_riskA) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value",
               values_transform = as.character) %>%
  knitr::kable()
```

## 2. Effect of phytosanitary measures (ENT2A vs ENT2B)

`ENT2A` asks how readily the pest could be transferred on a pathway **without**
official management measures; `ENT2B` asks the same **with** current measures in
place. Risk in scenario B should therefore be ≤ scenario A. If the two answers
are similar, measures have little modelled effect.

### 2.1 Risk with vs without measures

```{r p-risk-scatter, fig.height=5.5}
lim <- max(c(mgmt$riskA, mgmt$riskB), na.rm = TRUE)
ggplot(mgmt, aes(riskA, riskB)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = taxGroup), size = 2.4, alpha = 0.8) +
  coord_equal(xlim = c(0, lim), ylim = c(0, lim)) +
  labs(title = "Measures: no change for most species, large effect where they apply",
       subtitle = "Points on the dashed line = measures had no effect; points below = risk reduced",
       x = "Median RISK A (no measures)", y = "Median RISK B (with measures)",
       colour = "Taxonomic group") +
  theme(legend.position = "right", legend.text = element_text(size = 8))
```

## 3. Effect of import volume (ENT3 / "no import")

`ENT3` records the volume of host commodity imported: *Non-existent* (0),
*Small* (1), *Medium* (2), *Large* (3). For a group-1 pathway the entry score is
proportional to `ENT3`, so *Non-existent* import drives that pathway's entry — and
the resulting risk — to zero. Risk is shown for the actual scenario **B** (with
current measures); scenario A is overlaid to show how little the measures shift
risk at any import level.

### 3.1 Risk rises with import volume

```{r p-risk-by-import, fig.height=5}
species %>%
  filter(!is.na(max_import_cat)) %>%
  select(max_import_cat, `B (with measures)` = median_RISKB,
         `A (no measures)` = median_RISKA) %>%
  pivot_longer(c(`B (with measures)`, `A (no measures)`),
               names_to = "Scenario", values_to = "risk") %>%
  mutate(Scenario = factor(Scenario,
                           levels = c("B (with measures)", "A (no measures)"))) %>%
  ggplot(aes(max_import_cat, risk, fill = Scenario)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(0.8), width = 0.7) +
  scale_fill_manual(values = c("B (with measures)" = "#1f78b4",
                               "A (no measures)"   = "#a6cee3")) +
  labs(title = "Higher import volume, higher risk — measures barely shift it",
       subtitle = paste("Species grouped by their highest ENT3 import category.",
                        "B (dark) is the actual risk; A (light) the no-measure case."),
       x = "Maximum import volume (ENT3) across pathways",
       y = "Median RISK", fill = "Scenario")
```

## 4. Why is risk zero? Driver decomposition

Risk = `IMPACT × ENTRY × ESTABLISHMENT`. Any factor of zero forces risk to zero.
This uses the **actual** scenario **B** (with current measures): of
**`r n_zero_riskB`** species at median risk 0, the measures account for only
**`r n_pushed_to_zero`** (without measures `r n_zero_riskA` would still be zero).
For each zero-risk species we flag which component(s) are zero, so "no import" can
be compared against establishment and impact as the binding constraint.

```{r driver-decomp}
zero_risk <- species %>%
  filter(is_zero(median_RISKB)) %>%
  mutate(
    z_entry = is_zero(median_ENTRYB),
    z_estab = is_zero(median_ESTABLISHMENT),
    z_impact = is_zero(median_IMPACT),
    # within zero entry, is it driven by no import?
    entry_no_import = z_entry & zero_import %in% TRUE)

# count each binding component (a species can have several)
driver_counts <- tibble(
  Driver = c("Entry = 0", "  ↳ of which: all imports Non-existent",
             "Establishment = 0", "Impact = 0"),
  Species = c(sum(zero_risk$z_entry),
              sum(zero_risk$entry_no_import),
              sum(zero_risk$z_estab),
              sum(zero_risk$z_impact)))
knitr::kable(driver_counts,
             caption = paste0("Binding zero components among the ",
                              nrow(zero_risk),
                              " species at median risk 0 (components can co-occur)"))
```

```{r p-driver-bar, fig.height=4.5}
zero_risk %>%
  transmute(scientificName,
            `Entry = 0`         = z_entry,
            `Establishment = 0` = z_estab,
            `Impact = 0`        = z_impact) %>%
  pivot_longer(-scientificName, names_to = "Component", values_to = "is_zero") %>%
  filter(is_zero) %>%
  count(Component) %>%
  ggplot(aes(reorder(Component, n), n)) +
  geom_col(fill = "#41b6c4") +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  expand_limits(y = max(driver_counts$Species) * 1.1) +
  labs(title = "What forces risk to zero?",
       subtitle = paste0(nrow(zero_risk), " species at median risk 0 — ",
                         "components can co-occur"),
       x = NULL, y = "Species with this component = 0")
```

## 5. The two levers contrasted

```{r levers-contrast}
contrast <- tibble(
  Lever = c("Phytosanitary measures (ENT2A → ENT2B)",
            "No import (ENT3 = Non-existent)"),
  Mechanism = c("Reduces ENT2 by ~0–1 category; multiplies one factor of entry",
                "Sets the ENT3 factor of entry to 0 for group-1 pathways"),
  `Effect on risk` = c(
     paste0("No change for ", n_no_mgmt_effect, "/", n_species,
            " species; where answers differ, median ",
            scales::percent(median_rel_affected, accuracy = 1), " reduction"),
     paste0("Risk collapses to 0 (", n_zero_import,
            " zero-import species)")))
knitr::kable(contrast)
```

**Interpretation.** The assessors' observation is half right. In the stored
answers, official phytosanitary measures (`ENT2B`) very often equal the unmanaged
case (`ENT2A`): `r n_no_mgmt_effect` of `r n_species` species
(`r scales::percent(pct_no_effect, accuracy = 1)`) show no difference at all. But
where the answers *do* differ — typically by a single category — risk does not
move "only slightly": because `ENT2` enters the entry score multiplicatively, a
one-step drop (e.g. *Large* → *Medium*) roughly halves it, giving a median
reduction of `r scales::percent(median_rel_affected, accuracy = 1)` among affected
species. The practical lesson is that the perceived "small" gap between `ENT2A`
and `ENT2B` understates the modelled risk consequence. A *Non-existent* import
volume is different in kind: it removes a multiplicative factor of entry outright,
which is why "no import" drives risk to zero — although Section 4 shows that for
many species zero risk is actually set by establishment or impact, not by import
at all.

```{r session, include=FALSE}
try(dbDisconnect(con), silent = TRUE)
```

---

*Generated read-only from ``r params$db_path``. Methods: latest simulation per
assessment; medians from `simulationSummaries`; option→point maps parsed from the
question `list` JSON; all answer comparisons use the "likely" value.*
